DBES Grand Challenge : High throughput, scalable solution for processing taxi trip data

### 2015-03-26 paper/experiments.tex
fixing meanings

\section{Experiments}
We conducted several experiments to study the scalability of our solution. Firstly we study the throughput and mean message latency variation with the number of \textit{QueryEvaluators} for both single JVM and distributed setups. In an ideal system, we should observe linear throughput increment with a constant message latency. However there are some exceptions and we have provided our explanations on them. Finally we study the throughput variation with the window size to study the suitability of our solution for higher window sizes. All these experiments were performed on a LAN with a network bandwidth of 1Gbps and all machines were Intel(R) Xeon(R) 2.4GHz, 4 core duo machines with 16 GB of memory.

@@

Figure \ref{throughput_route} shows the throughput variation. For single JVM setup throughput increases from one instance to two instances. This is due to higher parallelism system achieves with two threads. After that throughput slightly decreases due to higher context switching with many threads. Similarly for distributed setup initially throughput increases with the number of   \textit{RouteProcessors} and achieves a steady limit. We observe two reasons for this non scalability after initial stages. Firstly, \textit{RouteProcessor} is not a computing intensive process. Secondly, there is only one \textit{RouteEventEmitter} and the speed at which it can read and send messages becomes the bottleneck. Initial throughput for distributed setup is less, due to message serialization, message deserialization and tcp overheads.

@@

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{profictgraph.png}
        \caption{Process graph to evaluate top ten profitable cells.}
        \label{profictgraph}
\end{figure}

@@

Figure \ref{throughput_profit} shows the throughput variation for both setups. For single JVM set up initially throughput increases, due to increment of parallelism and comes to a steady state. For distributed setup  throughput increases linearly, although it has an initial lesser throughput compared to single JVM setup due to message serialization, message deserialization and tcp overheads.

>>>>

\section{Experiments}
We conducted several experiments to study the scalability of our solution. Firstly we study the throughput and mean message latency variation with the number of \textit{QueryEvaluators} for both single JVM and distributed setups to find the scalability of the parallel evaluation. In an ideal system, we should observe linear throughput increment with a constant message latency. However there are some exceptions and we have provided our explanations on them. Finally we study the throughput variation with the window size to study the suitability of our solution for higher window sizes. All these experiments were performed on a LAN with a network bandwidth of 1Gbps and all machines were Intel(R) Xeon(R) 2.4GHz, 4 core duo machines with 16 GB of memory.

@@

Figure \ref{throughput_route} shows the throughput variation. For single JVM setup throughput increases from one instance to two instances. This is due to higher parallelism system achieves with two threads. After that throughput slightly decreases due to higher context switching with many threads. Similarly for distributed setup initially throughput increases with the number of   \textit{RouteProcessors} and achieves a steady limit. We observe two reasons for this non scalability after initial stages. Firstly, \textit{RouteProcessor} is not a computing intensive process. Secondly, there is only one \textit{RouteEventEmitter} and the speed at which it can read and send messages becomes the bottleneck. Initial throughput for distributed setup is less, due to message serialization, message deserialization and TCP overheads.

@@

\begin{figure}[!t]
        \centering
        \includegraphics[width=3.0in]{profitgraph.png}
        \caption{Process graph to evaluate top ten profitable cells.}
        \label{profictgraph}
\end{figure}

@@

Figure \ref{throughput_profit} shows the throughput variation for both setups. For single JVM set up initially throughput increases, due to increment of parallelism and comes to a steady state. For distributed setup  throughput increases linearly, although it has an initial lesser throughput compared to single JVM setup due to message serialization, message deserialization and TCP overheads.

### 2015-03-26 paper/query_evaluation.tex
fixing meanings

In this section, we explain how to evaluate queries with the given dataset. For this query evaluation we can use a distributed event stream processing system with the process graph as shown in Figure \ref{twonode_graph} for both queries.

@@

For frequent route query, route events can be partitioned using the hash code of the route object and process different route sets in different \textit{QueryEvaluators}. In such a setup, each \textit{QueryEvaluator} generates an event when the top ten routes get changed for its route set. Since this is not the desired result, we can add another process to aggregate these top result events and generate global route change events. Figure \ref{parallel} shows the possible execution patterns with the new process.

>>>>

In this section, we explain how to evaluate queries with the given data set. For this query evaluation we can use a distributed event stream processing system with the process graph as shown in Figure \ref{twonode_graph} for both queries.

@@

For frequent route query, route events can be partitioned using the hash code of the route object and process different route sets in different \textit{QueryEvaluators}. In such a setup, each \textit{QueryEvaluator} generates an event when the top ten routes get changed for its route set. Since this is not the desired result, we can add another process to aggregate these top result events and generate global route change events. Figure \ref{parallel} shows, possible parallel evaluation techniques with the new process. Multiple \textit{QueryEvaluator} instances can be executed in a single JVM using multiple threads or in a distributed setup with many JVMs using TCP to communicate.

### 2015-03-26 paper/query_implementation.tex
fixing meanings

@@

The frequent route query entails finding the most frequent routes within the last 30 minutes. Algorithm \ref{frequent_route_algorithm} shows the algorithm to process this query using the \textit{NodeList} data structure and a queue to keep the events for the time window.  Since queue operations are supported in O(1) time, query evaluations happen in O(log(n)) time. Algorithm \ref{frequent_route_algorithm} shows the basic algorithm to process an event. This algorithm assumes there is a global \textit{NodeList} object and route counts are stored in an Object called \textit{RouteCount}.

@@

We have implemented profitable cells query in a similar manner to the frequent route query using  time windows. Complexity of the query raised following issues and we have solved them as given below.
\begin{enumerate}
	\item The profitability is calculated using the median profit and the number of empty taxis. In order to calculate median with O(log(n)) time complexity, we use two heaps (\textit{java.util.PriorityQueue}) . One is used to keep the values lesser than the median value (max heap) and other is to keep higher values (min heap). However the \texttt{delete} operation of \textit{java.util.PriorityQueue} takes O(n) time, since it traverse along the array to find the index of the object. We implemented a similar algorithm using our \textit{DynamicHeap} to keep higher and lower values than median with value count. In worst case it takes O(log(n)) time to add or remove a value, in best case (i.e if the value is already present and that does not cause median to change) it takes O(1) time.  But that did not improved the performance. First, these heap sizes do not grow over 145 and there are same values repeating in the list increasing the probability of finding a value without traversing the whole list. Secondly, our implementation create more objects compared to just keeping double values though the time complexity is O(log(n)). 
	\item In order to keep track of empty taxis, we need to process both pickup cell and dropoff cell details for each message. One way of doing this is to send all events for the same process instance and process both details with one event. However this avoids the possibility of partitioning data across different processes. We generate two events, one for pickup data and one for dropoff data to make the process parallel. 
	\item In order to handle expiration of dropoff events, we keep track of the taxi ids currently available in a particular cell. If a pickup event occurs before dropoff event expiration, then we remove that taxi id from that cell and reduce the empty taxi count. In such a case, dropoff event expiration does not cause a reduction of taxi count.
\end{enumerate}

>>>>

\subsection{Finding Runtime Median}
In general, we can find runtime median using two heaps. One to keep values greater than the current median and one to keep values less than the current median. An implement of this algorithm using two \textit{java.util.PriorityQueue}s have an O(log(n)) time complexity for add operation and O(n) time complexity for remove operation (\texttt{delete} operation of \textit{java.util.PriorityQueue} takes O(n) time).  We overcame this issue using \textit{DynamicHeap} to keep values. In our solution, we use value as the key to insert and delete from the \textit{DynamicHeap} and keep value with its count (same value can be repeated) in the \textit{DynamicHeap}. In worst case it has O(log(n)) time complexity for both add and remove operations and in best case (i.e value already present and that does not cause median change) it has O(1) time complexity.

@@

The frequent route query entails finding the most frequent routes within the last 30 minutes. Algorithm \ref{frequent_route_algorithm} shows the algorithm to process this query using the \textit{NodeList} data structure and a queue to keep the events for the time window.  Since queue operations are supported in O(1) time, query evaluations happen in O(log(n)) time. Algorithm \ref{frequent_route_algorithm} shows the basic algorithm to process an event. This algorithm assumes there is a global \textit{NodeList} object and route counts are stored in an object called \textit{RouteCount}.

@@

We have implemented profitable cells query in a similar manner to the frequent route query using  time windows. Complexity of the query raised following issues and we have solved them as given below.
\begin{enumerate}
	\item The profitability is calculated using the median profit and the number of empty taxis. In order to calculate median, we use \textit{java.util.PriorityQueue} based approach mentioned in (Section 2.4). We choose this option over our implementation, because this approach performs better with smaller window sizes. First, these heap sizes do not grow over 145 and there are same values repeating in the list increasing the probability of finding a value without traversing the whole list. Secondly, our implementation create more objects per operation compared to just keeping double values though the time complexity is O(log(n)). In section 4.3 we show that our approach works better for larger window sizes.
	\item In order to keep track of empty taxis, we need to process both pickup cell and dropoff cell details for each message. One way of doing this is to send all events for the same process instance and process both details with one event. However this avoids the possibility of partitioning data across different processes. We generate two events, one for pickup data and one for dropoff data to make the process parallel. 
	\item In order to handle expiration of dropoff events, we keep track of the taxi ids currently available in a particular cell. If a pickup event occurs before dropoff event expiration, then we remove that taxi id from that cell and reduce the empty taxi count. In such a case, dropoff event expiration does not cause a reduction of taxi count.
\end{enumerate}
